\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2022-05-10 manuscript/paper.tex
update algorithm 3

Consider an independent and identically distributed (i.i.d.) cohort of $n$ subjects
\begin{equation}
  \label{eq:dataset-def}
  \cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\}
\end{equation}
where for each subject $i=1, \ldots, n$, process $Y_i^l$ is measured $n_i^l$ times at $t_{i1}^l, \ldots, t_{in_i^l}^l$ (which can differ between subjects and outcomes) with $t_{ij}^l \leq t_{ij+1}^l$ for all $j=1, \ldots, n_i^l-1$ and such that
\[ y_i^l=(y_{i1}^l, \ldots, y_{in_i^l}^l)^\top \in \R^{n_i^l} \quad \text{ with } \quad y_{ij}^l=Y_i^l(t_{ij}^l) \]
for all $l=1, \ldots, L$. 
For the $i$-th subject, let us denote 
\[
\left\{
    \begin{array}{ll}
        y_i &= ({y_i^1}^\top \cdots {y_i^L}^\top)^\top \in \R^{n_i},\\
        b_i &= ({b_i^1}^\top \cdots {b_i^L}^\top)^\top \in \R^r,
    \end{array}
\right.
\]
with $n_i = \sum_{l=1}^L n_i^l$ and $r = \sum_{l=1}^L r_l$ the total number of longitudinal measurements (for subject $i$) and the total dimension of the random effects respectively, as well as the following design matrices
\[ U_i = 
\begin{bmatrix}
  U_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & U_{iL}
\end{bmatrix} 
\in \R^{n_i \times q}
\qquad \text{and} \qquad
V_i = 
\begin{bmatrix}
  V_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & V_{iL}
\end{bmatrix}
\in \R^{n_i \times r}
\]
with $q = \sum_{l=1}^L q_l$ and where for all $l=1, \ldots, L$, one writes
\[
\left\{
    \begin{array}{ll}
        U_{il} &= \big(u_i^l(t_{i1}^l)^\top \cdots u_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times q_l},\\
        V_{il} &= \big(v_i^l(t_{i1}^l)^\top \cdots v_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times r_l}.
    \end{array}
\right.
\]
From now on, all computations are done conditionally on the design matrices $(U_i)_{i=1, \dots, n}$ and $(V_i)_{i=1, \dots, n}$.
Denoting $(t_1^u, \ldots, t_J^u)$ the $J \in \N_+$ unique failure times, we then extract all representation vectors $\big(\Psi_i(t_j^u)\big)_{j=1,\ldots,J}$ for all subjects $i=1, \dots, n$ before inference and from now on, all computations are also done conditionally on them.
Now, let us make two first hypothesis.
\begin{assumption}
\label{assumption1}
Suppose that the random effects are independent of the latent class membership, and that the latter remain independent conditional on the observed data (namely $T$, $\Delta$ and $Y$).
\end{assumption}
Assumption~\ref{assumption1} states that subject-and-longitudinal outcome specific random effects $b^l$ do not depend on the latent class membership, which is not a strong modeling assumption.
\begin{assumption}
\label{indep-hyp-2}
Suppose that for a given subject $i$ and conditional on the random effects and the latent class membership, all $Y_i^l$ are independent for $l=1, \dots, L$.
\end{assumption}
Assumption~\ref{indep-hyp-2} is similar to standard modeling hypothesis, see for instance~\citet{tsiatis2004joint}.
For all $k = 0, \ldots, K-1$, we denote 
\[ \beta_k = ({\beta_k^1}^\top \cdots {\beta_k^L}^\top)^\top \in \R^q \] 
and 
\[M_{ik} = U_i\beta_k + V_ib_i \in \R^{n_i}. \]

@@

\paragraph{Procedure.}
Let us describe here the procedure we follow to evaluate model performances on simulated data in Section~\ref{sec:simulation study} and on real data in Section~\ref{sec:application}.
\begin{algorithm}[!htb]
\DontPrintSemicolon
\caption{Procedure followed to assess model performances in our real-time prediction paradigm.}
\label{alg:evaluation}
\KwInput{Dataset $\cD_n$~\eqref{eq:dataset-def} \textbf{only if} \textit{\texttt{simulation = false}} ; a \texttt{model} under study}
\tcc{We run $\breve{R}=100$ independant experiments}
$seed_0 = 0$ \\
\For{$\breve{r} = 1,\ldots,\breve{R}$}
{
  $seed_{\breve{r}} = seed_{\breve{r}-1} + 1$ \\
  \If{\texttt{simulation = true}}
  {
    {$\cD_n \leftarrow \texttt{simulate}(seed_{\breve{r}})$ \tcc{Sample a new dataset at each round $\breve{r}$}}
  }
  $(\cD_{n_{train}}, \cD_{n_{test}}) \leftarrow \texttt{split\_train\_test}(\cD_n, seed_{\breve{r}})$ \\
  \tcc{For simplicity, we do not detail the cross-validation procedure here}
  $\hat \theta \leftarrow$ \texttt{model.fit}$(\cD_{n_{train}})$ \\
  \For{$i = 1,\ldots,n_{test}$}
  {
    $t_i^{max} \sim t_i \times \big(1 - \text{Beta}(2, 5)\big)$ \tcc{See Figure~\ref{fig:timax-pdf}}
    \tcc{Truncate $y_i$}
    $\mathcal{X}_i = (x_i, y_i)$ \\
    \If{\normalfont \texttt{model = lights}}
    {
      $\big(\Psi_i(t_j^u)\big)_{j=1,\ldots,J} \leftarrow \texttt{extract\_features}(y_i)$ \tcc{$(t_j^u)_{j=1,\ldots,J}$ from $\cD_{n_{train}}$}
      $\mathcal{X}_i = \Big(x_i, y_i, \big(\Psi_i(t_j^u)\big)_{j=1,\ldots,J}\Big)$ 
    }
    $\hat \cR_{i,\breve{r}} \leftarrow \texttt{model.predict}(\mathcal{X}_i)$
  }
}
\textbf{Return:} {$\big[\hat \cR_{i,\breve{r}}\big]_{\substack{i=1,\ldots,n_{test} \\ \breve{r}=1,\ldots,\breve{R} \ \ \ }}\ , \ \big[\texttt{time}_{i,\breve{r}}\big]_{\substack{i=1,\ldots,n_{test} \\ \breve{r}=1,\ldots,\breve{R} \ \ \ }}$}
\end{algorithm}

>>>>

Consider an independent and identically distributed (i.i.d.) cohort of $n$ subjects
\begin{equation}
  \label{eq:dataset-def}
  \cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\}
\end{equation}
where for each subject $i=1, \ldots, n$, process $Y_i^l$ is measured $n_i^l$ times at $t_{i1}^l, \ldots, t_{in_i^l}^l$ (which can differ between subjects and outcomes) with $t_{ij}^l \leq t_{ij+1}^l$ for all $j=1, \ldots, n_i^l-1$ and such that
\begin{equation}
  \label{eq:def-yil}
  y_i^l=(y_{i1}^l, \ldots, y_{in_i^l}^l)^\top \in \R^{n_i^l} \quad \text{ with } \quad y_{ij}^l=Y_i^l(t_{ij}^l) 
\end{equation}
for all $l=1, \ldots, L$. 
For the $i$-th subject, let us denote 
\[
\left\{
    \begin{array}{ll}
        y_i &= ({y_i^1}^\top \cdots {y_i^L}^\top)^\top \in \R^{n_i},\\
        b_i &= ({b_i^1}^\top \cdots {b_i^L}^\top)^\top \in \R^r,
    \end{array}
\right.
\]
with $n_i = \sum_{l=1}^L n_i^l$ and $r = \sum_{l=1}^L r_l$ the total number of longitudinal measurements (for subject $i$) and the total dimension of the random effects respectively, as well as the following design matrices
\[ U_i = 
\begin{bmatrix}
  U_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & U_{iL}
\end{bmatrix} 
\in \R^{n_i \times q}
\qquad \text{and} \qquad
V_i = 
\begin{bmatrix}
  V_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & V_{iL}
\end{bmatrix}
\in \R^{n_i \times r}
\]
with $q = \sum_{l=1}^L q_l$ and where for all $l=1, \ldots, L$, one writes
\[
\left\{
    \begin{array}{ll}
        U_{il} &= \big(u_i^l(t_{i1}^l)^\top \cdots u_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times q_l},\\
        V_{il} &= \big(v_i^l(t_{i1}^l)^\top \cdots v_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times r_l}.
    \end{array}
\right.
\]
From now on, all computations are done conditionally on the design matrices $(U_i)_{i=1, \dots, n}$ and $(V_i)_{i=1, \dots, n}$.
Denoting $(t_1^u, \ldots, t_J^u)$ the $J \in \N_+$ unique failure times, we then extract all representation vectors $\big(\Psi_i(t_j^u)\big)_{j=1,\ldots,J}$ for all subjects $i=1, \dots, n$ before inference and from now on, all computations are also done conditionally on them.
Now, let us make two first hypothesis.
\begin{assumption}
\label{assumption1}
Suppose that the random effects are independent of the latent class membership, and that the latter remain independent conditional on the observed data (namely $T$, $\Delta$ and $Y$).
\end{assumption}
Assumption~\ref{assumption1} states that subject-and-longitudinal outcome specific random effects $b^l$ do not depend on the latent class membership, which is not a strong modeling assumption.
\begin{assumption}
\label{indep-hyp-2}
Suppose that for a given subject $i$ and conditional on the random effects and the latent class membership, all $Y_i^l$ are independent for $l=1, \dots, L$.
\end{assumption}
Assumption~\ref{indep-hyp-2} is similar to standard modeling hypothesis, see for instance~\citet{tsiatis2004joint}.
For all $k = 0, \ldots, K-1$, we denote 
\[ \beta_k = ({\beta_k^1}^\top \cdots {\beta_k^L}^\top)^\top \in \R^q \] 
and 
\[M_{ik} = U_i\beta_k + V_ib_i \in \R^{n_i}. \]

@@

\paragraph{Procedure.}
Let us describe now in Algorithm~\ref{alg:evaluation} the procedure we follow to evaluate model performances on simulated data in Section~\ref{sec:simulation study} and on real data in Section~\ref{sec:application}. Here we do not give details on the cross-validation step, but it follows the same idea : for each fold, we sample the $t_i^{max}$ and truncate $y_i$ for $i$ in the test set of the fold.
\begin{algorithm}[!htb]
\DontPrintSemicolon
\caption{Procedure followed to assess performances of a given model in our real-time prediction paradigm.}
\label{alg:evaluation}
\KwInput{Dataset $\cD_n$~\eqref{eq:dataset-def} \textbf{only if} \textit{\texttt{simulation = false}} ; a \texttt{model} under study}
\KwOutput{Confidence intervals on C-index metric as well as on running time.}
\tcc{We run $\breve{R}=100$ independant experiments}
$seed_0 = 0$ \\
\For{$\breve{r} = 1,\ldots,\breve{R}$}
{
  $\texttt{start\_time} = \texttt{time}()$ \\
  $seed_{\breve{r}} = seed_{\breve{r}-1} + 1$ \\
  \If{\texttt{simulation = true}}
  {
    {$\cD_n \leftarrow \texttt{simulate}(seed_{\breve{r}})$ \tcc{Sample a new dataset at each round $\breve{r}$}}
  }
  $(\cD_{n_{train}}, \cD_{n_{test}}) \leftarrow \texttt{split\_train\_test}(\cD_n, seed_{\breve{r}})$ \\
  $\texttt{model.cross\_validate}(\cD_{n_{train}})$ \tcc{Select the best hyper parameters}
  \texttt{model.fit}$(\cD_{n_{train}})$ \tcc{Learn $\hat \theta$}
  \For{$i = 1,\ldots,n_{test}$}
  {
    $t_i^{max} \sim t_i \times \big(1 - \text{Beta}(2, 5)\big)$ \tcc{See Figure~\ref{fig:timax-pdf}}
    $y_i \leftarrow \big(Y_i^l(t_{ij}^l) \big)_{\substack{j=1, \ldots, n_i^l-1 \\ l=1, \ldots, L \ \ \ }} \text{ with } t_{ij}^l \leq t_i^{max}$ \tcc{Truncate $y_i$, see~\eqref{eq:def-yil}}
    $\mathcal{X}_i = (x_i, y_i)$ \\
    \If{\normalfont \texttt{model = lights}}
    {
      $\big(\Psi_i(t_j^u)\big)_{j=1,\ldots,J} \leftarrow \texttt{extract\_features}(y_i)$ \tcc{$(t_j^u)_{j=1,\ldots,J}$ from $\cD_{n_{train}}$}
      $\mathcal{X}_i = \Big(x_i, y_i, \big(\Psi_i(t_j^u)\big)_{j=1,\ldots,J}\Big)$ 
    }
    $\hat \cR_{i,\breve{r}} \leftarrow \texttt{model.predict}(\mathcal{X}_i)$
  }
  $\texttt{score}_{\breve{r}} \leftarrow \texttt{c\_index}\big((\hat \cR_{i,\breve{r}}, t_i, \delta_i)_{i=1,\ldots,n_{test}} \big)$ \\
  $\texttt{end\_time} = \texttt{time}()$ \\
  $\texttt{running\_time}_{\breve{r}} = \texttt{end\_time} - \texttt{start\_time}$ \\
}
\textbf{Return:} {$\big[\texttt{score}_{\breve{r}}\big]_{\breve{r}=1,\ldots,\breve{R}}\ , \ \big[\texttt{running\_time}_{\breve{r}}\big]_{\breve{r}=1,\ldots,\breve{R}}$}
\end{algorithm}
